---
title: "R Udacity Bikeshare Project"
output: html_notebook
---

```{r}
#load necessary libraries 
library(ggplot2)
library(tidyr)
library(dplyr)
library(stringr)
library(reshape)

```
```{r}
#load data in a separate chunk 
ny = read.csv('new_york_city.csv')
wash = read.csv('washington.csv')
chi = read.csv('chicago.csv')
```

```{r}

```
```{r}

#Question 1:
#Is there a drop off in usage in the winter 

#convert pull month out of start.time.
nyMonth = months(as.Date(ny$Start.Time))
washMonth = months(as.Date(wash$Start.Time))
chiMonth  = months(as.Date(chi$Start.Time))
#gather the data by month
nyMonthCounts   = as.data.frame(table(nyMonth))
washMonthCounts = as.data.frame(table(washMonth))
chiMonthCounts  = as.data.frame(table(chiMonth))

#create a single dataframe with all of the counts by month for the different cities
monthCounts <- cbind(nyMonthCounts, washMonthCounts$Freq, chiMonthCounts$Freq)
#need to add a value to the table to sort easily by month for correct
#ordering of the histogram bars on the plot
values <- c("1", "2", "3", "4", "5", "6")
index <- c("January", "February", "March", "April","May","June")
monthCounts$date_col <- values[match(monthCounts$nyMonth, index)]
monthCounts <- monthCounts[order(monthCounts$date_col),]
colnames(monthCounts) = c("month","New York","Washington DC","Chicago","date_col")

df <- data.frame(month = monthCounts$date_col,
                  NY = monthCounts$`New York`,
                  Washington = monthCounts$`Washington DC`,
                  Chicago = monthCounts$Chicago
                )


dfMut <- df %>% gather(key = Month, value = Value, NY:Chicago)

ggplot(dfMut, aes(Month, Value, fill = month)) + 
  geom_col(position = "dodge") + 
  scale_fill_discrete(name = "Month", labels = c("Jan", "Feb","Mar","Apr","May","Jun")) +
  labs(title="Number of rides per month for each city") +
  xlab("City and Month") +
  ylab("Number of Rides")
  

```
In all three cities there is a clear trend of lower usuage in the winter months.   Without the full year of data over multiple years, it is hard to make a definitive determination as to whether or not usage is lower in the winter.   But, from the data that exists, there is a definite trend towards having lower usage in the winter months.


```{r}
#Question 2:
#Rank start location to find out which are the most popular
#stations and on average how long are the trips taken from those stations.  This has
#implications for making sure there are enough bikes in the correct location
#as well as determining how many bikes need to be recovered
#also how many total usages where there per city.


#gets the statistics of the data  and take the top 30 station data in terms
# of the most used start stations for plotting
getStats <- function(df1, df2){
 
rowDim = dim(nyStationSum)[1]
retDF <- data.frame(matrix(nrow=rowDim, ncol=5))
colnames(retDF) <- c("Start.Station","Avg_time","std_time","median","Freq")
index = 0

for (i in df1$Var1){
       data <- df2[df2$Start.Station == i,]
       rowData <- df1[df1$Var1 == i,]
      
       retDF[index,1] = str_wrap(i, width = 15)
       retDF[index,2] = mean(data$Trip.Duration)
       retDF[index,3] = sd(data$Trip.Duration)
       retDF[index,4] = median(data$Trip.Duration)
       retDF[index,5] = rowData$Freq
       index = index + 1
  
}

  
  retDF <- retDF[order(-retDF$"Freq"),]
  retDF <- head(retDF, 10)
  return (retDF)
}

#get the Station and trip duration for each city
nyStart <- as.data.frame(ny$Start.Station)
nyStart <- cbind(nyStart, ny$Trip.Duration)
colnames(nyStart) = c("Start.Station","Trip.Duration")
washStart <- as.data.frame(wash$Start.Station)
washStart <- cbind(washStart,wash$Trip.Duration)
colnames(washStart) = c("Start.Station","Trip.Duration")
chiStart  <- as.data.frame(chi$Start.Station)
chiStart <- cbind(chiStart,chi$Trip.Duration)
colnames(chiStart) = c("Start.Station","Trip.Duration")


#summarize data by start.station
nyStationSum   <- as.data.frame(table(ny$Start.Station))
washStationSum <- as.data.frame(table(wash$Start.Station))
chiStationSum  <- as.data.frame(table(chi$Start.Station))

totalNY   = sum(nyStationSum$Freq) - 1
totalWash = sum(washStationSum$Freq) - 1
totalChi  = sum(chiStationSum$Freq)  -1
total_usage = totalNY + totalWash + totalChi
x = c(totalNY,totalWash,totalChi)
#get percentage usage per city
x = 100*(x/total_usage)
city = c("New York", "Washington DC", "Chicago")
dfx = data.frame(values = x)
dfx <- cbind(dfx, city)
ggplot(aes(city,values), data = dfx) + 
  geom_col() + 
  labs(title = "Percent Total bike usages per city") +
  xlab("City") + 
  ylab("Percent")
#calculate statistics on trip duration for each of these start stations and
#take only the top 10 stations in terms of number of times the station was used
 
statNY   <- getStats(nyStationSum, nyStart)
statWash <- getStats(washStationSum, washStart)
statChi  <- getStats(chiStationSum, chiStart)


#create plots 
ggplot(aes(Start.Station, Freq), data = statNY) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(title = "New York top 10 start locations by number of usages") + 
  xlab("Start Location") + 
  ylab("Number of usages")

ggplot(aes(Start.Station, Freq), data = statWash) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(title = "Washington DC top 10 start locations by number of usages") + 
  xlab("Start Location") + 
  ylab("Number of usages")

ggplot(aes(Start.Station, Freq), data = statWash) + 
  geom_col() + 
  theme(axis.text.x = element_text(angle = 90)) + 
  labs(title = "Chicago top 10 start locations by number of usages") + 
  xlab("Start Location") + 
  ylab("Number of usages")

```






```{r}
#Question 3
#Which group, customers, subscribers or none, have more usage and does 
#that very by city?
getCustomerInfo <- function(city,df){
  #print(city)
  #print(colnames(df))
  #print(dim(df[df$User.Type == "Subscriber",]))
  total = dim(df)[1]
  numSubscribers = as.integer(100*(dim(df[df$User.Type == "Subscriber",])[1])/total)
  numCustomers   = as.integer(100*(dim(df[df$User.Type == "Customer",])[1])/total)
  numNone        = as.integer(100*(dim(df[(df$User.Type != "Customer" & df$User.Type != "Subscriber"),])[1])/total)
 
  
  totalVal = c(city,numSubscribers, numCustomers, numNone)

return(totalVal)
  
}


customerDF <- data.frame(matrix(nrow=3, ncol=4))
colnames(customerDF) <- c("City","Subscribers","Customers","None")
customerDF[1,] = getCustomerInfo("New York", ny)
customerDF[2,] = getCustomerInfo("Washington CD", wash)
customerDF[3,] = getCustomerInfo("Chicago", chi)


#use line plot to plot the percentage data.   
ggplot( customerDF) +
  geom_line(aes(City, as.integer(Subscribers),color = "Subscribers", group = 1)) +
  geom_line(aes(City, as.integer(Customers),color = "Customers", group = 1)) +
  labs(title = "Percentage of User Types by City") +
  xlab("City") + 
  ylab("Percentage") +
  coord_cartesian( ylim = c(0, 100))
```
The majority of users are subscribers, topping out at close to 90% in New York.  There are some data points where the type of user was not included and those data points have been excluded.  



